Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-05-24 analysis/paper/paper.Rmd
Draft to coauthors.

---
title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false

institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

Analysis methods that can help make better use of available monitoring data to inform decisions include Bayesian population modelling [@schaub_integrated_2021] and value of information analyses [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012]. To be useful, these methods must be tailored to particular species and decision contexts. Here, we offer a Bayesian population model for boreal caribou (_Rangifer tarandus caribou_) that integrates information about the impacts of landscape change with available demographic data to reduce uncertainty in population viability projections for boreal caribou [building on @eacker_web-based_2019].  Following the value of information analysis framework proposed by Dunham et al. (2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. To help make the approach more accessible, we offer an open source app that allows users to analyze existing local demographic data, or to explore alternative monitoring scenarios with simulated data. Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.

Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016]. Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020]. The causes of this relationship include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011]. Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].

@@

Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020]. Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. In Bayesian population models, all available demographic or population data (e.g. abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021]. However Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships or composition survey errors.

Here, we present a Bayesian population model for boreal caribou [building on @eacker_web-based_2019] designed to reduce uncertainty in projections of the impacts of changing landscape conditions by integrating prior information from national analysis of demographic-disturbance relationships with available local demographic data. Information is shared among populations using priors instead of integrating multiple populations into a hierarchical analysis [@leasure_hierarchical_2019] because obtaining permission to use caribou demographic data can be difficult, and most managers only have access to their own data. This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships. We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and status assessments, improve viability projections on changing landscapes, and help clarify where, why and when additional data is required to inform decisions and project outcomes on changing landscapes. In cases where available information is sufficient for assessing and projecting population viability, there is a risk of misallocating limited resources to monitoring that could be better spent on conservation action.

While we believe that using available information to reduce uncertainty is important, we also believe it is important to acknowledge remaining uncertainty. In general, parametric Bayesian population models include a number of simplifying assumptions, and misspecified models can yield biased or overconfident results [@schaub_integrated_2021]. More specifically, models that do not account for observation errors can be misleading [@schaub_integrated_2021]. Several authors have pointed out that  misclassification of adults and missing calves in composition surveys can bias recruitment estimates from those surveys [@ellington_divergent_2020; @decesare_estimating_2012], but none have thus far examined sensitivity of Bayesian caribou population models to these sources of error [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019]. We address this gap with a new bias correction term derived from a simple model of the recruitment survey observation process.

@@

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
two-stage demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) a delayed reproduction term to account for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum potential recruitment of female calves per cow $\dot{X}_t$:
\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
is the probability of misidentifying young bulls as adult females and
vice versa, and $\dot{z}$ is the probability of missing a calf. We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.

@@

## Integration of local demographic data and national disturbance-demographic relationships in a Bayesian population model

@@

Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the _survival_ R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
  (\#eq:hatSEst)
\end{equation}

As in the simulations (eq \@ref(eq:barS)), we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
[@johnson_science_2020; @stewart_climate-informed_2023; @dyson_existing_2022]. We add a Gaussian distributed random year effect to account for stochastic variation among years [@eacker_web-based_2019]:
\begin{equation}
S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
  (\#eq:S)
\end{equation}

@@

knitr::kable(tbl,caption="Prior means and standard errors of Gaussian distributed survival and reproduction model parameters. Standard errors are calculated by dividing Johnson et al's (2020) reported confidence intervals by 3.92. Multipliers on the standard errors (*) are calibrated so that the 95\\% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5\\% and 97.5\\% quantiles of 1000 simulated survival
and recruitment trajectories (Supplement C).")
#TO DO: indicate which of these are from Johnson, and which are calibrated.
```

As in the simulations, we assume a simple two-class demographic model, with a census in late
winter of surviving females and new recruits to the post-juvenile (aged
1+) class. The number of post-juvenile females that survive from year
$t$ to the census $W_t$ is binomially distributed with survival
probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The number of
juveniles recruiting to the post-juvenile class at the census is a
Poisson distributed function of the number of surviving post-juvenile
females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Poisson}(W_tX_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y=3$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 5000 animals. For reference, all
variables and parameters are defined in supplemental Table A1.

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
-->
We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the _R2jags_ R package [@su_r2jags_2021] to fit the Bayesian model. We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2. We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the _mcmcplots_ R package [@curtis_mcmcplots_2018].

@@

## User interface and open, flexible, reuseable implementation

Analysis was implemented using the R language [@r_core_team_r_2023] and a variety of _tidyverse} packages [@wickham_welcome_2019]. To make the work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios, borrowing and building on code from an app developed by Eacker et al. [-@eacker_web-based_2019] where appropriate. Our app allows users to provide existing local data, or to make projections informed by simulated data as we have done in this analysis (similar to Figure
\@ref(fig:examplePredictions)). The app allows advanced users to adjust many model parameters (supplemental Table A1). Default parameter values are set to allow novice users to get reasonable results quickly. Most of the code needed to run the app is in an R package, _caribouMetrics_, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g. @dyson_existing_2022; @stewart_climate-informed_2023]. The _caribouMetrics_ package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>. Find documentation and vignettes on the package website
<https://landscitech.github.io/caribouMetrics/>. The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>. Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.

@@

>>>>

---
title: "Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false

institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country. Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and reduce the risk that lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process.  We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where uncertainty is higher, deploying fewer collars for more years may be better than intensive short-term monitoring for improving the reliability of status projections. 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

Analysis methods that can help make better use of available monitoring data to inform decisions include Bayesian population modelling [@schaub_integrated_2021] and value of information analyses [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012]. To be useful, these methods must be tailored to particular species and decision contexts. Here, we offer a simple Bayesian population model for boreal caribou (_Rangifer tarandus caribou_) that integrates information about the impacts of landscape change with available demographic data to reduce uncertainty in population viability projections for boreal caribou [building on @eacker_web-based_2019].  Following the value of information analysis framework proposed by Dunham et al. (2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.

Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016]. Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020]. The causes of these demographic-disturbance relationships include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011]. Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].

Interventions such as wolf culls, maternal penning, and restoration of linear features are intended to improve survival and reproduction of populations that are not otherwise self-sustaining [@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a need to assess and compare effects of boreal caribou conservation actions with careful monitoring. Our focus in this paper is assessing and projecting self-sustainability of populations in the absence of these interventions, to clarify when additional information is needed to reduce uncertainty about whether there is a need for action. Where actions are taken, outcomes should be monitored; the simple model we propose here does not include effects of management interventions, and cannot be used to project outcomes for managed populations.

@@

Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020]. Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. In Bayesian population models, all available demographic or population data (e.g. abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021]. However, there is very little monitoring data available for some caribou populations [@johnson_science_2020] and there is a risk that lack of local data will be used as a reason to delay conservation action [@boan_climate_2018]. Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships, and we are concerned that models informed only by local demographic data could be used to inflate uncertainty about the status of poorly monitored populations.

We propose to reduce uncertainty in assessments and projections of local caribou population viability by integrating prior information from national analysis of demographic-disturbance relationships with available local demographic data in a simple Bayesian population model for boreal caribou [building on @eacker_web-based_2019]. In our simple proof-of-concept model, information is shared among populations using priors derived from a published analysis of variation in demographic rates among populations across Canada [@johnson_science_2020]. A more rigorous statistical approach would be to integrate multiple populations into a hierarchical Bayesian analysis of variation within and among populations [@leasure_hierarchical_2019]; we have not done this because owners have not authorized the use of the data for this purpose, obtaining permission to use caribou demographic data is difficult, and most managers only have access to their own local data. Despite limitations, our simple Bayesian model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships. We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and status assessments, improve viability projections on changing landscapes, and help clarify where, why and when additional data is required to inform decisions and project outcomes on changing landscapes. We hope that this approach can help clarify where available information is sufficient for assessing and projecting population viability, and thereby reduce the risk that uncertainty is used as a reason to delay conservation action for imperiled populations.

While we believe that using available information to reduce uncertainty is important, we also believe it is important to acknowledge remaining uncertainty. In general, parametric Bayesian population models include a number of simplifying assumptions, and misspecified models can yield biased or overconfident results [@schaub_integrated_2021]. More specifically, models that do not account for observation errors can be misleading [@schaub_integrated_2021]. Although several authors have pointed out that  misclassification of adults and missing calves in composition surveys can bias recruitment estimates from those surveys [@ellington_divergent_2020; @decesare_estimating_2012] none have thus far examined sensitivity of Bayesian caribou population models to these sources of error [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019]. We address this gap with a bias correction term derived from a simple model of the recruitment survey observation process.

@@

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) a delayed reproduction term to account for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum potential recruitment of female calves per cow $\dot{X}_t$:
\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
is the probability of misidentifying young bulls as adult females and
vice versa, and $\dot{z}$ is the probability of missing a calf. We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.

@@

## Bayesian integration of local demographic data and national disturbance-demographic relationships

@@

Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the _survival_ R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a modelled as a Gaussian
distributed function of the true survival probability, truncated between 0 and 1, to allow sharing survival information across years with a random effect [@eacker_web-based_2019]:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t)T(0,1).
  (\#eq:hatSEst)
\end{equation}

For consistency with [@johnson_science_2020], we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
[@stewart_climate-informed_2023; @dyson_existing_2022]. We add a Gaussian distributed random year effect to account for stochastic variation among years [@eacker_web-based_2019]:
\begin{equation}
S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
  (\#eq:S)
\end{equation}

@@

knitr::kable(tbl,caption="Prior means and standard errors of Gaussian distributed survival and reproduction model parameters. Standard errors are calibrated so that the 95\\% prior prediction intervals for survival and recruitment from the Bayesian model match the range between the 2.5\\% and 97.5\\% quantiles of 1000 simulated survival
and recruitment trajectories (Supplement C).")
#TO DO: indicate which of these are from Johnson, and which are calibrated.
```

As in the simulations, we assume a simple demographic model, with a census in late
winter of surviving females and new recruits to the post-juvenile (aged
1+) class. The number of post-juvenile females that survive from year
$t$ to the census $W_t$ is binomially distributed with survival
probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The number of
juveniles recruiting to the post-juvenile class at the census is a
Poisson distributed function of the number of surviving post-juvenile
females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Poisson}(W_tX_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y=3$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 5000 animals. For reference, all
variables and parameters are defined in supplemental Table A1.

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
-->

Models and analysis were implemented using the R language [@r_core_team_r_2023] and a variety of _tidyverse_ packages [@wickham_welcome_2019]. We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the _R2jags_ R package [@su_r2jags_2021] to fit the Bayesian model. We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2. We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the _mcmcplots_ R package [@curtis_mcmcplots_2018]. Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>, and in the  _caribouMetrics_ R package available on GitHub at <https://github.com/LandSciTech/caribouMetrics>.

@@

@@

## Bayesian integration of local demographic data & national demographic-disturbance relationships

The availability of boreal caribou demographic data varies widely across Canada due to large variation in habitat condition, population condition, available resources for monitoring, and remoteness. In a few regions enough data has been collected over time to characterize local demographic-disturbance relationships [@rudolph_demographic_2017;@hervieux_widespread_2013], but most populations have only been sparsely monitored for snapshot estimates of trend or abundance [@johnson_science_2020]. As a result, in most regions it is necessary to make assumptions informed by observed variation in demographic rates with disturbance among populations and other relevant information about impacts of disturbance on caribou. However, high observed variation in demography where anthropogenic disturbance is low leads to high uncertainty in projections informed by national demographic-disturbance relationships alone [@stewart_climate-informed_2023;@dyson_existing_2022]. In some cases, predictions of population growth rate informed by national demographic-disturbance relationships appear to disagree with estimates from local demographic data [@eccc_scientific_2011; @dyson_existing_2022], but without formal methods of accounting for uncertainty it can be difficult to distinguish true discrepancies from stochastic variation.

We developed a simple Bayesian population model [building on @eacker_web-based_2019] to help address these challenges by acknowledging uncertainty associated with local demographic data and national demographic-disturbance relationships, and integrating these two sources of information to reduce uncertainty in projections of the impacts of changing landscape conditions. This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships.

Our model assumes that increasing cumulative anthropogenic disturbance causes a log-linear decrease in caribou demographic rates (equations 4, 5, 10 and 13), reflecting a scientific consensus informed by a number of lines of evidence that cumulative anthropogenic disturbance has a negative impact on boreal caribou [@eccc_scientific_2011; @johnson_science_2020; @gouvernement_du_quebec_revue_2021]. Local data can be integrated to estimate local demographic-disturbance parameters that differ from the national means, and in the absence of local data, appropriate priors lead to a range of outcomes consistent with the observed range of variation across the country [Figures 3 and 5 in @johnson_science_2020]. Others [@moeller_integrating_2021; @mcnay_demographic_2022; @eacker_web-based_2019] have also used Bayesian models to estimate caribou demographic rates from local monitoring data, but our addition of prior information from national demographic-disturbance relationships is novel, and important for reducing the risk that lack of local data will be used as a reason to delay conservation action for highly disturbed and poorly monitored populations.  Integrating demographic-disturbance relationships into a viability projection model clarifies that no additional information is needed to project outcomes where disturbance is high and actions to support recovery are not taken.

There are a variety of possible actions intended to improve survival and reproduction of populations that are not otherwise self-sustaining [@mcnay_demographic_2022; @serrouya_predicting_2020]. The simple models in this paper do not include these actions, and cannot be used to project outcomes for managed populations. Our intention is to reduce uncertainty that can be used to delay conservation action; there is nothing in our arguments, models or results that can or should be used to discourage much needed monitoring of the effects of conservation actions. Even when no additional additional monitoring is required for assessing self-sustainability, there are a variety of reasons to continue monitoring highly disturbed boreal caribou populations. Our point is that action is also required, and lack of monitoring should not be used as a reason to delay conservation measures for highly disturbed populations.

Variation in anthropogenic disturbance across the Canadian boreal region is partially correlated with variation in climate and primary productivity because forestry occurs in higher productivity areas, and petroleum extraction activities are concentrated in Western Canada; concerns have been raised about confounding of demographic-disturbance relationships [@wilson_causal_2021;@neufeld_primary_2021; @neilson_direct_2022]. One option for addressing these concerns would be to develop regional demographic-disturbance relationships for populations that are ecologically similar to one another (e.g. caribou inhabiting forests that are sufficiently productive for industrial forestry in Eastern Canada) [@eccc_scientific_2024], and use the priors from regional demographic-disturbance relationships to inform demographic projections within the region. However, while acknowledging and allowing for regional variation, we also need to guard against risks that uncertainty will be inflated and used as an excuse to delay action, a tactic that has been used in this and other decision contexts [@bissonnette_controverses_2020; @boan_climate_2018]. In order to be informative and credible, regional demographic-disturbance analyses must include a sufficient number of observations spanning a range from high to low anthropogenic disturbance. Priors that lead to projections of outcomes for caribou that fall outside the observed range of variation across the country, characterized here (Figure 2) by 95% predictive intervals from a national analysis (Johnson et al. 2020), are implausible. In other words, if we have seen no examples of self-sustaining caribou populations at a particularly high level of disturbance, it is unreasonable to project or expect this outcome. These premises constrain the plausible demographic-disturbance priors, and clarify that no additional information is needed to predict outcomes for caribou where anthropogenic disturbance is high (Figure 5) and no meaningful actions are taken to support recovery.

## Allowing for composition survey errors

Another novel aspect of this work is a term that allows for bias associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012]. Lacking information about the true magnitude of this bias, we examined sensitivity to a range of assumed values, and found that adding unknown bias did increase uncertainty, but these increases were not large (supplement E). A key reason for this result is that we assumed the same range of values for the probability of misclassifying adults and the probability of missing calves. Presuming that surveyed groups contain more cows than bulls, and that there is an equal probability of misclassifying cows and young bulls and vice versa, then misclassification errors cause high estimated recruitment. Missing calves causes low estimated recruitment, and together these errors tend to cancel one another out. We did not investigate sensitivity to the assumption that new recruits are 50% females, but we note that this is another possible source of error that would most likely lead to low recruitment estimates.

Given lack of knowledge about true error rates, the most appropriate interpretation of our results is that better information on composition survey error rates is required in order to assess their importance. Fecal pellets are proving to be an interesting source of information on sex and age ratios in surveyed groups [@mcfarlane_genetic_2018;@jones_novel_2023]; for example, in analysis of 14 populations of boreal and southern mountain caribou, Jones et al. 2023 found that sampled groups contained more females than males, and that mothers tend to stay with their daughters and sons, while fathers do not. These results are broadly consistent with our assumption that composition survey groups include mothers and younger animals of both sexes that can be difficult to classify visually. Documenting animals that are difficult to classify as “unknown” can help analysts minimize and account for classification bias [@ellington_divergent_2020], as can comparing classification results from multiple independent observers. The probability of missing calves is more difficult to quantify and account for using aerial survey methods alone, but likely increases with vegetation cover.

Alternatives to aerial composition surveys for estimating recruitment include fecal pellet analysis [@mcfarlane_genetic_2018; @mcfarlane_spatial_2021;@mcfarlane_identification_2022] and calf collaring [@ellington_divergent_2020]. Each method has costs and benefits, and the reality is that monitoring everything everywhere all the time in great detail is neither feasible nor necessary. We hope that a bias correction term for the simple recruitment-mortality model [@hatter_moose_1991;@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020] that is used for many herd viability assessments will enable and encourage practitioners to examine the sensitivity of their conclusions to plausible composition survey error rates, and help clarify thinking about where, when, and why additional information is needed.

## Limitations and Future Possibilities
